--- 
title: "Advanced Visualisation and Data Wrangling in R"
author: "TJ McKinley"
site: bookdown::bookdown_site
output:
    bookdown::gitbook:
        config:
            sharing: null
        css: 'style.css'
        includes:
            in_header: _toggle.html
    bookdown::pdf_book:
        keep_tex: yes
        includes:
            in_header: header.tex
linkcolor: blue
documentclass: book
link-citations: yes
---

# Introduction {-}

```{r, child = "_setup.Rmd", include = F, purl = F, cache = F}
```

This workshop requires that you're comfortable with R, and specifically with the concept of `data.frame` objects. The ability to work with and visualise data frames is one of the key reasons why R is so popular amongst statisticians and data scientists. Although a vast amount can be achieved using base R functionality, one of R's other key strengths is the vast array of [packages](https://cran.r-project.org/web/packages/) that it supports, which add a rich variety of additional functionality to R.

A suite of packages that are fast becoming *de rigueur* for performing myriad data science tasks is known as the [`tidyverse`](https://www.tidyverse.org/). These packages provide powerful functions for doing visualisation and manipulation of complex data sets. In this workshop we will introduce key `tidyverse` packages, such as `readr`, `tidyr`, `dplyr` and `ggplot2`, and show how they can be used to efficiently process and visualise complex data.

## `tidyverse` packages {-}

The [`tidyverse`](https://www.tidyverse.org/) is a suite of packages, including `tidyr`, `dplyr`, `ggplot2`, `purrr`, `tibble` and `readr`. Although these packages can each be installed and loaded separately, they are designed to work together, and as such will simply install and load the `tidyverse` directly, rather than worry too much about which functions belong to which packages.

To install `tidyverse`, use:

```{r, eval = F}
install.packages("tidyverse")
```

and once installed, it can be loaded using:

```{r, eval = F}
library(tidyverse)
```

in the usual way.

> **Note**: if you are loading `tidyverse` as part of an R Markdown document, and you want to knit to a PDF document using LaTeX, then it sometimes throws an error when loading because LaTeX can't process the correct fonts for the loading message. Hence in R Markdown documents I always suppress the load messages through the chunk option `message = F` e.g.
> ````
> ```{r, message = F}`r ''`
> library(tidyverse)
> ```
> ```` 

## Data {-}

All data files can be downloaded as a ZIP file from [here](https://exeter-data-analytics.github.io/AdVis/dataFiles.zip).

## Tasks {-}

```{task}
All **tasks** will be denoted in panel boxes like this one. In the HTML version, all solutions can be toggled by hitting the **Show Solution** buttons. In the PDF version solutions are given in the Appendix and are linked via the **Show Solution** buttons.
```

```{solution}
The answer is, of course, 42.
```
