--- 
title: "NERC DTP Visualisation Workshop"
author: "TJ McKinley"
site: bookdown::bookdown_site
output:
    bookdown::pdf_book:
        includes:
            in_header: header.tex
    bookdown::gitbook:
        config:
            sharing: null
            toc:
                before: |
                    <a href="https://www.exeter.ac.uk/"><img src="_colour_logo_vector.svg" width="280"></a>
        css: 'style.css'
        includes:
            in_header: _toggle.html
documentclass: book
link-citations: yes
description: "NERC DTP Visualisation Workshop---Lecture Notes"
---

# Introduction {-}

```{r, child = "_setup.Rmd", include = FALSE, purl = FALSE, cache = FALSE}
```

```{r include = FALSE, purl = FALSE, cache = FALSE}
## set seed for reproducibility
set.seed(123654)
#opts_chunk$set(cache = TRUE)
```

This workshop will introduce you to using the [`ggplot2`](https://ggplot2.tidyverse.org/) package in R to quickly and easily produce visualisations of (complex) data sets. This is part of a wider suite of packages known as the [`tidyverse`](https://www.tidyverse.org/), which I can thoroughly recommend.

## Recommended reading {-}

The following books are probably the most relevant to this workshop:

* [R for Data Science](https://www.oreilly.com/library/view/r-for-data/9781491910382/) by Hadley Wickham and Garrett Grolemund (2016), available for free online at [http://r4ds.had.co.nz/](http://r4ds.had.co.nz/);
* [ggplot2](https://www.springer.com/gp/book/9780387981413) by Hadley Wickham (2009), available for free online at [https://ggplot2-book.org/](https://ggplot2-book.org/).

## Structure of this workshop {-}

R can be tricky to learn, so please feel free to e-mail me with questions at [t.mckinley@exeter.ac.uk](mailto:t.mckinley@exeter.ac.uk), though please make sure that you have thought through your question in advance. Most R problems can be solved with judicious use of [Google](https://www.google.com/) and [StackOverFlow](https://stackoverflow.com/)...

In this workshop, R commands to be entered into the console are shown in grey boxes e.g.

```{r, eval = FALSE}
seq(0, 1, length = 11)
```

and the corresponding outputs look like:

```{r, echo = FALSE}
seq(0, 1, length = 11)
```

If you're reading this online, there is a downloadable PDF version of these notes if you prefer (available via the link next to the title at the top of the page).

## Tasks {-}

```{task}
All **tasks** will be denoted in panel boxes like this one. In the HTML version, all solutions can be toggled by hitting the **Show Solution** buttons. In the PDF version solutions are given in the Appendix and are linked via the **Show Solution** buttons.
```

```{solution}
The answer is, of course, 42. 
```

