--- 
title: "Introduction to R"
author: "TJ McKinley"
site: bookdown::bookdown_site
output:
    bookdown::gitbook:
        config:
            sharing: null
        css: 'style.css'
        includes:
            in_header: _toggle.html
    bookdown::pdf_book:
        keep_tex: yes
        includes:
            in_header: header.tex
linkcolor: blue
documentclass: book
link-citations: yes
description: "Introduction to R Tutorial"
---

# Acknowledgements

```{r, child = "_setup.Rmd", include = F, purl = F, cache = F}
```

```{r include = FALSE, purl = F, cache = F}
## set seed for reproducibility
set.seed(123654)
```

I am indebted to [Dr Matt Castle](http://www.plantsci.cam.ac.uk/directory/castle-matt), whose excellent Introduction to R practicals form the basis of the first part of these notes. I heartily recommend the following books:
    
* [Statistics: An Introduction Using R](https://www.amazon.com/Statistics-Introduction-Michael-J-Crawley/dp/0470022981/ref=sr_1_1?s=books&ie=UTF8&qid=1484432671&sr=1-1&keywords=an+introduction+to+statistics+with+r+crawley) by Mick Crawley (2005);
* [The R Book](https://www.amazon.com/R-Book-Michael-J-Crawley/dp/0470973927/ref=sr_1_1?s=books&ie=UTF8&qid=1484432596&sr=1-1&keywords=the+r+book) by Mick Crawley (2007).

[Hadley Wickham](http://hadley.nz/) and colleagues have also written many great R packages and several great books:
    
* [Advanced R](https://www.amazon.com/dp/1466586966/ref=cm_sw_su_dp?tag=devtools-20) by Hadley Wickham (2014), which has a useful companion [website](http://adv-r.had.co.nz/); and
* [R for Data Science](https://www.amazon.com/R-Data-Science-Hadley-Wickham/dp/1491910399/ref=as_li_ss_tl?ie=UTF8&qid=1469550189&sr=8-1&keywords=R+for+data+science&linkCode=sl1&tag=devtools-20&linkId=6fe0069f9605cf847ed96c191f4e84dd) by Hadley Wickham and Garrett Grolemund (2016), also with a great companion [website](http://r4ds.had.co.nz/).

Finally, [ggplot2](https://www.amazon.com/dp/0387981403/ref=cm_sw_su_dp?tag=ggplot2-20) by Hadley Wickham (2009) is also great for learning the [ggplot2](http://ggplot2.org/) graphics package in R.
