# (PART) Advanced visualisation and Data Wrangling in R {-}

# Introduction to `tidyverse`

Now that you're comfortable with R, and specifically with the concept of `data.frame` objects, we can begin to really start unlocking the power of R for analysing complex data sets. 

The ability to work with and visualise data frames is one of the key reasons why R is so popular among statisticians and data scientists. Although a vast amount can be achieved using base R functionality, one of R's other key strengths is the vast array of [packages](https://cran.r-project.org/web/packages/) that it supports, which add a rich variety of additional functionality to R.

A suite of packages that are fast becoming *de rigueur* for performing myriad data science tasks is known as the [`tidyverse`](https://www.tidyverse.org/). These packages provide powerful functions for doing visualisation and manipulation of complex data sets. In this workshop we will introduce key `tidyverse` packages, and show how they can be used to efficiently process and visualise complex data sets.

## `tidyverse` packages {-}

The [`tidyverse`](https://www.tidyverse.org/) is a suite of packages, including e.g. `tidyr`, `dplyr`, `ggplot2`, `purrr`, `tibble` and `readr`. Although these packages can each be installed and loaded separately, they are designed to work together, and as such will simply install and load the `tidyverse` directly, rather than worry too much about which functions belong to which packages.

As usual, if you want to install `tidyverse`, use:

```{r, eval = FALSE}
install.packages("tidyverse")
```

and once installed, it can be loaded using:

```{r, eval = FALSE}
library(tidyverse)
```

in the usual way.

## Data files and slides {-}

All data files can be downloaded as a ZIP file from [here](https://exeter-data-analytics.github.io/AdVis/dataFiles.zip). PDF copies of the workshop slides can be found through the following links:

* [Advanced Visualisation](advis/AdVisSlides/AdVisHandout.pdf)
* [Data Wrangling](advis/DataWrSlides/DataWrHandout.pdf)
