---
title: "Sexual activity in fruitflies"
author: TJ McKinley
output: html_document
---

```{r, include = F, purl = F}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A cost of increased reproduction in terms of reduced longevity has been shown for female fruitflies, but not for males. This experiment used a factorial design to assess whether increased sexual activity affected the lifespan of male fruitflies.

The flies used were an outbred stock. Sexual activity was manipulated by supplying individual males with one or eight receptive virgin females per day. The longevity of these males was compared with that of two control types. The first control consisted of two sets of individual males kept with one or eight newly inseminated females. Newly inseminated females will not usually remate for at least two days, and thus served as a control for any effect of competition with the male for food or space. The second control was a set of individual males kept with no females. There were 25 males in each of the five groups, which were treated identically in number of anaesthetizations (using CO2) and provision of fresh food medium.

The data should have the following columns:

* **id**: a ID for each fly in each group (1--25).
* **partners**: number of companions (0, 1 or 8).
* **type**: type of companion (0: newly pregnant female; 1: virgin female; 9: not applicable (when 'partners = 0')).
* **longevity**: lifespan, in days.
* **thorax**: length of thorax, in mm.
* **sleep**: percentage of each day spent sleeping.
* **group**: experimental group (1: control; 2: 1 newly pregnant partner; 3: 8 newly pregnant partners; 4: 1 virgin partner; 5: 8 virgin partners).

Source: [Partridge and Farquhar (1981)](http://www.annualreviews.org/doi/pdf/10.1146/annurev.pu.04.050183.001103)

# Analysis

First we load and summarise the data set:

```{r}
## load the data
ff <- readRDS("ff.rds")

## summarise data
summary(ff)
```

Now plot `thorax` against `group`: 

```{r}
## plot thorax against group
plot(thorax ~ group, data = ff, 
     ylab = "Length of thorax (mm)", 
     xlab = "Experimental group")
```

The plot suggests that thorax length does not seem to vary very much between the experimental groups. Now we plot `longevity` against `group`:

```{r}
## plot longevity against group
plot(longevity ~ group, data = ff,
     ylab = "Longevity (days)", 
     xlab = "Experimental group")
```

The plot suggests that longevity does not seem to vary very much between the experimental groups, with the exception of group 5 (and possibly group 4), which exhibits lower longevity. This is the group where males have 8 virgin sexual partners (i.e. the group that corresponds to a higher degree of sexual reproduction). Finally we plot `longevity` against `thorax`:

```{r}
## plot longevity against thorax with points stratified
## by experimental group
plot(longevity ~ thorax, data = ff, 
     pch = 20, 
     xlab = "Length of thorax (mm)",
     ylab = "Longevity (days)")
legend(0.65, 95, legend = paste("Group", 1:5), pch = 20)
```

We can see that there is a strong linear relationship between `thorax` and `longevity`, such that larger flies tend to live longer. On top of this there may be some additional impact of the experimental group, but we will need to investigate further to fully quantify these effects.
