---
title: "Lecture Title"
author: "T.J. McKinley ([t.mckinley@exeter.ac.uk](mailto:t.mckinley@exeter.ac.uk))"
output: 
  revealjs::revealjs_presentation:
    css: ../setupFiles/_styles.css
    reveal_options:
      width: 1200
      height: 900
      slideNumber: true
      previewLinks: true
    self_contained: false
    reveal_plugins: ["chalkboard", "zoom"]
    include:
      in_header: ../setupFiles/_script.html
  beamer_presentation:
      latex_engine: xelatex
      includes:
          in_header: ../setupFiles/_header.tex
---

```{r, setup, child = "../setupFiles/_setup.Rmd", include = FALSE, purl = FALSE, cache = FALSE}
```

## Some mathematics

Some mathematics:

\begin{align*}
    \frac{dS(t)}{dt} &= - \frac{\beta S(t)I(t)}{N(t)}\\
    \frac{dI(t)}{dt} &=  \frac{\beta S(t)I(t)}{N(t)}
\end{align*}

## Multiple columns

`r bcols("t")`
`r bcol(0.48)`

Column content 1

$$
    x = y^2
$$
  
`r ecol()`
`r bcol(0.48)`

Column content 2

$$
    x = z + y^3
$$

`r ecol()`
`r ecols()`

## Multiple columns

Columns can also be centre aligned with the `center` argument, and can also change width using the `width` argument.

`r bcols()`
`r bcol(0.68)`

Column content 1 70% width

$$
    x = y^2
$$

`r ecol()`
`r bcol(0.28, center = TRUE)`

Column content 2 30% width (centred)

$x = z + y^3$

`r ecol()`
`r ecols()`

## Code

```{r, echo = TRUE}
head(iris)
```

You can also change the code/output size using `size = "smaller"` or `"larger"`:

```{r, echo = TRUE, size = "smaller"}
head(iris)
```

## Footnotes

Footnotes can also be added by placing an inline R chunk containing `footnotemark()` at the point in the slide where you want the footnote mark added, and then a custom `footnotes` chunk at the end of the slide code containing the footnote text e.g.

`r benv("smaller")`
`r ifelse(opts_knit$get("rmarkdown.pandoc.to") != "beamer", "<code>Here is a footnote&#96;r footnotemark()&#96;</code>", "\\texttt{Here is a footnote\\textasciigrave r footnotemark()\\textasciigrave}")`
`r eenv("smaller")`

````
```{footnotes}`r ''`
some **markdown** text here
```
````

Here is a footnote`r footnotemark()`

```{footnotes}
some **markdown** text here
```

## Footnotes

Multiple footnotes can use multiple `footnotmark()` calls and then the footnote text separated by `####` in the footnotes chunk e.g.

`r benv("smaller")`
`r ifelse(opts_knit$get("rmarkdown.pandoc.to") != "beamer", "<code>Here is a footnote&#96;r footnotemark()&#96; and another footnote&#96;r footnotemark()&#96;</code>", "\\texttt{Here is a footnote\\textasciigrave r footnotemark()\\textasciigrave and another footnote\\textasciigrave r footnotemark()\\textasciigrave}")`
`r eenv("smaller")`
````
```{footnotes}`r ''`
footnote **one** text here
####
footnote **two** text here
```
````
Here is a footnote`r footnotemark()` and another footnote`r footnotemark()`
`r br()`

```{footnotes}
footnote **one** text here
####
footnote **two** text here
```

## Miscellaneous

Boxes can be added using a `box` chunk e.g.

````
```{box}`r ''`
Some **important** boxed text
```
````

```{box}
Some **important** boxed text
```

The `wrap = TRUE` chunk option can also be provided for beamer slides if text wrapping required.

## Environments

Environments like `smaller`, `larger` and `center` can be added using e.g. `benv("smaller")` and `eenv("smaller")` in R chunks e.g.

`r benv("smaller")`

This is small

`r eenv("smaller")`

`r benv("center")`

This is centred!

`r eenv("center")`
